An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-26 scagnostics.Rmd
done with criteria description

keywords:
- tour
- projection pursuit 
- statistical graphics
- data visualisation
- exploratory data analysis
- high-dimensional data

@@

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers (\cite{diaconis84}, \cite{huber85}, \cite{jones87}, \cite{posse95b}). The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimisation feasible, small angles between projections correspond to small perturbations of the sample, and thus shold be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the crteria upon which the scagnostic indexes, and the several alternative indexes are examined:

  - smoothness: the index function should evolve smoothly between nearby projections, i.e. on an interpolated path we expect to observe slow and consistent change in the index function. \textcolor{red}{maybe can relate this to consistency consition developed by P. Hall, but don't understand this at the moment. In addition this relates to questions of optimiser choice and efficiency.} We evaluate smoothness both for transitions between noise variables and when moving from noise variables to an informative view in Section \ref{sec:smooth}.
  - distribution of index values: the distribution of index values when studying data points without structure should have a low variance, and be clearly separated from index values obtained when looking at structured data distributions. \textcolor{red}{I dont really understand this part. I think it can instead be related to the concept of squint angle \cite{barnett1981interpreting}.} We show examples in Section \ref{sec:dist}
  - ability to find fine and broad structure: \textcolor{red}{there are two different considerations, one  that relates to the concept of squint angle discussed in \cite{posse95b}. On the other hand should also consider detection of "fine" structure vs noise, this would be controlled e.g. by the allowed number of bins for the MINE index functions, additionally this will impact computing times, i.e. trade accuracy for efficiency}. We address these questions in Section \ref{sec:params} with an illustrative example.
  - rotation invariance: in the tour there is an assumed degeneracy between all projections obtained when rotating within the projected space only, for 2-d projections it means that we cannot assing x- and y- axis without ambiguity. A projection pursuit index should therefore be invariant under rotation of the basis in the projection plane. By costruction this is not the case for most index functions defined above. We illustrate the impact of non invariance in Section \ref{sec:rot}. \textcolor{red}{consider frame interpolation rather than plane interpolation, which will avoid issues of rotation when interpolating, reamining issue would be optimisation efficiency.}
  - speed: the computation time is an important factor in particular when the guided tour is used to view the optimisation in real time. Note that longer computation times may be acceptable when first recording the optimisation and replaying the guided tour once the maximum has been identified. Computation times for some benchmark scenarios are compared in Section \ref{sec:timer}.

@@

## Summary

Our results can be summarised by comparing

- scagnostics indices:
    - advantages: scaling with larger datasets thanks to binning, set of index functions covering different types of interesting structures
    - disadvantage: noisy, requires good optimisation algorithm, stringy too noisy to be used as index
- splines2D:
    - advantages: fast to compute, also considering scaling with larger data sets, efficient in detecting functional relations, typically smooth behaviour (but we have seen some exceptions)
    - disadvantage: not general, highly not rotation invariant
- dcor2D:
    - advantages: smoothly changing, sensitive to general structure
    - disadvantage: slow if computed without binning, large sensitivity to noise compared to MIC/TIC (example: low values obtained for the circle view of the pipe), sensitive to rotation
    - NOTE: I have not tested this much, should probably add binned version to timing plot
- MINE (MIC, TIC):
    - advantages: general, good discrimination even with noisy structures
    - disadvantages: slow to compute, noisy for small datasets, TIC sensitive to rotation

\begin{table}
\begin{center}
\caption{Summary}
\label{tab:summary}
\begin{tabular}{|l|c|c|c|c|}
\hline
Index & smooth & rotation invariant & fast & general \\
\hline
Convex & ! & ! & \checkmark & ! \\
Skinny & ! & ! & \checkmark & ! \\
Stringy & $\times$ & ! & \checkmark & ! \\
\hline
splines2D & \checkmark & $\times$ & \checkmark & ! \\
dcor2D & \checkmark & $\times$ & ! & ! \\
\hline
MINE & ! & ! & ! & \checkmark\\
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

keywords:
- statistical graphics
- data visualisation
- exploratory data analysis
- data science 
- guided tour

@@

XXX Need to add something here about the guided tour

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers (\cite{diaconis84}, \cite{huber85}, \cite{jones87}, \cite{posse95b}, \cite{hall89}). The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimisation feasible, small angles between projections correspond to small perturbations of the sample, and thus shold be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the crteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

- **smoothness**: This is the consistency property mentioned above. The index function values are examined over interpolated tour paths, that is, the value is plotted against time, where time indexes the sequence of projections produced by the tour path. The signature of a good PPI is that the plotted function is relatively smooth. The interpolation path corresponds to small angle changes between projections, so the value should be very similar. 
  should evolve smoothly between nearby projections, i.e. on an interpolated path we expect to observe slow and consistent change in the index function. \textcolor{red}{maybe can relate this to consistency consition developed by P. Hall, but don't understand this at the moment. In addition this relates to questions of optimiser choice and efficiency.} We evaluate smoothness both for transitions between noise variables and when moving from noise variables to an informative view in Section \ref{sec:smooth}.
- **squintability**: \cite{barnett1981interpreting} introduced the idea of squint angle to indicate resolution of structure in the data. Fine structure like the parallel planes in the infamous RANDU data \ref{randu} has a small squint angle because you have to be very close the the optimal projection plane to be able to see the structure. Structures with small squint angle are difficult to find, because the optmisation algorithm needs to get very close to begin hill-climbing to the optimum. The analyst doesn't have control over the data structure, but does have control over the PPI. Squintability is about the shape of the PPI over all projections. It should have smooth low values for noise projections and a clearly larger value, ideally with a big squint angle, for structured projections. The optimiser should be able to clearly see the optimal projections as different from noise. To examine squintability, the PPI values are examined on interpolated tour paths between a noise projection and a distant structured projection. 
distribution of index values: the distribution of index values when studying data points without structure should have a low variance, and be clearly separated from index values obtained when looking at structured data distributions. \textcolor{red}{I dont really understand this part. I think it can instead be related to the concept of squint angle \cite{barnett1981interpreting}.} We show examples in Section \ref{sec:dist}
- **flexibility**: An analyst can have a toolbox of indices that may cover the range of fine and broad structure, which underlies the scagnostics suite. Early indexes, based on density estimation could be programmed to detect fine or large structure by varying the binwidth. This is examined by using a range of structure in the simulated data examples. 
ability to find fine and broad structure: \textcolor{red}{there are two different considerations, one  that relates to the concept of squint angle discussed in \cite{posse95b}. On the other hand should also consider detection of "fine" structure vs noise, this would be controlled e.g. by the allowed number of bins for the MINE index functions, additionally this will impact computing times, i.e. trade accuracy for efficiency}. We address these questions in Section \ref{sec:params} with an illustrative example.-->
- **rotation invariance**: The orientation of structure within a projection plane should not change the index value. This is especially important when using the projection pursuit guided tour, because the tour path is defined between planes, along a geodesic path, not bases within planes. If a particular orientation is more optimal, this will get lost as the projection shown pays no attention to orientation. \cite{bujaXXX} describes alternative interpolation paths based on Givens and Householder rotations which progress from basis to basis. It may be possible to ignore rotation invariance with these interpolations but there isn't a current implementation, primarily because the within-plane spin that is generated is distracting from a visualisation perspective. Rotation invariance is checked for the proposed PPIs by rotating the structured projection, within the plane. <!--
in the tour there is an assumed degeneracy between all projections obtained when rotating within the projected space only, for 2-d projections it means that we cannot assing x- and y- axis without ambiguity. A projection pursuit index should therefore be invariant under rotation of the basis in the projection plane. By construction this is not the case for most index functions defined above. We illustrate the impact of non invariance in Section \ref{sec:rot}. \textcolor{red}{consider frame interpolation rather than plane interpolation, which will avoid issues of rotation when interpolating, reamining issue would be optimisation efficiency.}
- **speed**: Being fast to compute allows the index to be used in real-time in a guided tour, where the optimisation can be watched. When the computations are shifted off-line, to watch in replay, computation times matter less. This is checked  by comparing times for benchmark scenarios with varying sample size and dimension.

@@
